---
title: "Tables and Supplementary information"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r chunk-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  autodep = TRUE,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r wall-time-start, cache=FALSE, include=FALSE}
# Track time spent on performing this analysis
start_time <- Sys.time()
```

## Introduction
Here we generate tables and supplementary tables/information.

## Libraries and sources
These libraries and sources are used for this analysis.
```{r libs-and-src, message=FALSE, warning=FALSE, cache=FALSE}
library(tidyverse)
library(tidylog)
library(here)
library(openxlsx)
```

Definition of global variables that are used throughout this analysis.
```{r analysis-specific-params, cache=FALSE}
# keys to annotate contrasts
key_mm <- readRDS(here("data/meta-chronic-vs-acute/contrast_annotation.rds"))
key_hs <- readRDS(here("data/meta-mouse-vs-human/contrast_annotation.rds"))
```

## Differential expressed genes
### Mouse - Supplementary Table 1
```{r degs-mouse}
keys <- key_mm %>%
  distinct(contrast, label2, treatment, class)

contrasts <- readRDS(here("output/meta-chronic-vs-acute/limma_result.rds")) %>%
  filter(regulation != "ns") %>%
  select(-treatment, -source, -class) %>%
  inner_join(keys) %>%
  select(-contrast)

l <- contrasts %>%
  mutate(sheet_name = str_c(treatment, " (", class, ")")) %>%
  select(gene,
    contrast = label2, logFC, `t-statistic` = statistic,
    `p-value` = pval, fdr, regulation, sheet_name
  ) %>%
  arrange(contrast, `p-value`) %>%
  nest(data = -sheet_name) %>%
  deframe()

write.xlsx(l, here("tables/Supplementary Table 1.xlsx"),
  keepNA = TRUE
)
```

### Human - Supplementary Table 4
```{r degs-human}
keys <- key_hs %>%
  distinct(contrast, disease, author2, source, phenotype)

contrasts <- readRDS(here("output/meta-mouse-vs-human/limma_result.rds")) %>%
  filter(regulation != "ns") %>%
  inner_join(keys) %>%
  select(-contrast)

l <- contrasts %>%
  select(gene,
    contrast = disease, logFC, `t-statistic` = statistic,
    `p-value` = pval, fdr, regulation, sheet_name = author2
  ) %>%
  arrange(contrast, `p-value`) %>%
  nest(data = -sheet_name) %>%
  deframe()

# add etiology gene sets
etiology_gene_sets <- readRDS(here(
  "output/meta-mouse-vs-human/etiology_gene_sets.rds"
)) %>%
  arrange(etiology)
l$`Etiology gene sets` <- etiology_gene_sets

write.xlsx(l, here("tables/Supplementary Table xy human_degs.xlsx"),
  keepNA = TRUE
)
```

## Time-series cluster
### Gene lists - Supplementary table 2
```{r "gene-list-of-time-series-cluster"}
keys <- key_mm %>%
  drop_na(time) %>%
  distinct(treatment, time = value, time_label2, class)


profile_label <- tribble(
  ~profile, ~process, ~treatment, ~class,
  "STEM ID: 14", "Inflammation", "CCl4", "Chronic",
  "STEM ID: 6", "Proliferation", "CCl4", "chronic",
  "STEM ID: 17", NA_character_, "CCl4", "Chronic",
  "STEM ID: 12", NA_character_, "CCl4", "Chronic",
  "STEM ID: 7", "Metabolism (1)", "CCl4", "Chronic",
  "STEM ID: 9", "Metabolism (2)", "CCl4", "Chronic",
  "STEM ID: 13", "ECM", "CCl4", "Chronic",
  "STEM ID: 14", "Stress response", "APAP", "Acute",
  "STEM ID: 9", "Migration", "APAP", "Acute",
  "STEM ID: 12", "Proliferation", "APAP", "Acute",
  "STEM ID: 6", "Metabolism (1)", "APAP", "Acute",
  "STEM ID: 11", "Metabolism (2)", "APAP", "Acute",
  "STEM ID: 18", "Metabolism (3)", "APAP", "Acute",
  "STEM ID: 8", "Metabolism (4)", "APAP", "Acute"
)

stem_profiles <- list.files(here("output"), "stem_result.rds",
  recursive = T,
  full.names = TRUE
) %>%
  map_dfr(readRDS) %>%
  mutate(
    treatment = case_when(
      key == "apap" ~ "APAP",
      key == "bdl" ~ "Bile Duct Ligation",
      str_detect(key, "ccl4") ~ "CCl4",
      key == "hepatec" ~ "Partial Hepatectomy"
    ),
    class = case_when(
      key == "pure_ccl4" ~ "Chronic",
      TRUE ~ "Acute"
    )
  )

df <- stem_profiles %>%
  filter(p <= 0.05) %>%
  mutate(profile = str_c("STEM ID: ", as.character(profile))) %>%
  # match stem profiles with biological processes
  left_join(profile_label, by = c("treatment", "class", "profile")) %>%
  # match time points with nice labels
  left_join(keys) %>%
  mutate(time_label2 = as.character(time_label2)) %>%
  mutate(
    time_label2 = replace_na(time_label2, "0"),
    time_label2 = fct_inorder(str_c(time_label2, " (logFC)")),
    sheet_name = str_c(treatment, " (", class, ")")
  ) %>%
  # spread table
  select(gene, time_label2, value, profile, process, size,
    `p-value` = p,
    sheet_name
  ) %>%
  nest(data = -sheet_name) %>%
  transmute(sheet_name, table = data %>% map(function(data) {
    data %>%
      mutate(profile = fct_reorder(profile, `p-value`)) %>%
      spread(time_label2, value) %>%
      arrange(profile)
  })) %>%
  deframe()

write.xlsx(df,
  here("tables/Supplementary Table xy stem_results.xlsx"),
  keepNA = TRUE
)
```

## Exclusive gene sets - Supplementary Table 3
```{r exclusive-gene-sets}
exclusive_chronic <- readRDS(here(
  "output/meta-chronic-vs-acute/ranked_exclusive_chronic_genes.rds"
)) %>%
  mutate(sheet_name = "Exclusive chronic")

exclusive_acute <- readRDS(here(
  "output/meta-chronic-vs-acute/ranked_exclusive_acute_genes.rds"
)) %>%
  mutate(sheet_name = "Exclusive acute")

common <- readRDS(here(
  "output/meta-chronic-vs-acute/ranked_common_genes.rds"
)) %>%
  mutate(sheet_name = "Common")

df <- bind_rows(exclusive_chronic, exclusive_acute, common) %>%
  mutate(regulation = case_when(
    sheet_name == "Exclusive chronic" & chronic_statistic >= 0 ~ "up",
    sheet_name == "Exclusive chronic" & chronic_statistic < 0 ~ "down",
    sheet_name == "Common" & chronic_statistic >= 0 ~ "up",
    sheet_name == "Common" & chronic_statistic < 0 ~ "down",
    sheet_name == "Exclusive acute" & acute_statistic >= 0 ~ "up",
    sheet_name == "Exclusive acute" & acute_statistic < 0 ~ "down",
  )) %>%
  select(rank, gene, regulation, sheet_name) %>%
  nest(data = -sheet_name) %>%
  deframe()

write.xlsx(
  df,
  here("tables/Supplementary Table xy exclusive_common_genes.xlsx"),
  keepNA = TRUE
)
```

## Human mouse consistent genes
```{r human-moust-consistent-genes}
consistent_genes <- readRDS(here(
  "output/meta-mouse-vs-human/consistent_genes.rds"
))

t <- consistent_genes %>%
  distinct(gene, rank, contrast, logFC, celltype) %>%
  group_by(gene) %>%
  mutate(
    mean_sign = mean(sign(logFC)),
    regulation = case_when(
      mean_sign < 0 ~ "down",
      mean_sign >= 0 ~ "up"
    )
  ) %>%
  distinct(gene, rank, celltype, regulation) %>%
  group_by(gene, rank, regulation) %>%
  summarise(celltype = str_c(celltype, collapse = "|")) %>%
  ungroup() %>%
  arrange(rank) %>%
  select(rank, gene, regulation, celltype)


write.xlsx(
  t,
  here("tables/Supplementary Table xy consistent_genes.xlsx"),
  keepNA = TRUE
)
```

```{r wall-time-end, cache=FALSE, include=FALSE}
duration <- abs(as.numeric(difftime(Sys.time(), start_time, units = "secs")))
t <- print(sprintf("%02d:%02d", duration %% 3600 %/% 60, duration %% 60 %/% 1))
```
Time spend to execute this analysis: `r t` minutes.
