---
title: "Acute LPS mouse model"
author: "Christian H. Holland"
date: "2020-12-19"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r chunk-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  autodep = TRUE,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction
Here we analysis a mouse model of LPS induced acute liver damage. 

## Libraries and sources
These libraries and sources are used for this analysis.
```{r libs-and-src, message=FALSE, warning=FALSE, cache=FALSE}
library(mouse4302.db)

library(tidyverse)
library(tidylog)
library(here)

library(oligo)
library(annotate)
library(limma)
library(biobroom)

library(janitor)

library(AachenColorPalette)
library(cowplot)
library(lemon)
library(patchwork)

options("tidylog.display" = list(print))
source(here("code/utils-microarray.R"))
source(here("code/utils-utils.R"))
source(here("code/utils-plots.R"))
```

Definition of global variables that are used throughout this analysis.
```{r analysis-specific-params, cache=FALSE}
# i/o
data_path <- "data/mouse-acute-lps"
output_path <- "output/mouse-acute-lps"

# graphical parameters
# fontsize
fz <- 9
```

## Data processing
### Load .CEL files and quality control
The array quality is controlled based on the relative log expression values (RLE) and the normalized unscaled standard errors (NUSE).
```{r load-cel-files}
# load cel files and check quality
platforms <- readRDS(here("data/annotation/platforms.rds"))
raw_eset <- list.celfiles(here(data_path), listGzipped = T, full.names = T) %>%
  read.celfiles() %>%
  ma_qc()
```

### Normalization and probe annotation
Probe intensities are normalized with the `rma()` function. Probes are annotated with MGI symbols.
```{r normalization-and-annotation}
eset <- rma(raw_eset)

# annotate microarray probes with mgi symbols
expr <- ma_annotate(eset, platforms)
colnames(expr) = str_remove_all(colnames(expr), "_\\(Mouse430_2\\).CEL")

# save normalized expression
saveRDS(expr, here(output_path, "normalized_expression.rds"))
```

### Build meta data
Meta information are parsed from the sample names.
```{r build-meta-data}
# build meta data
meta <- colnames(expr) %>%
  enframe(name = NULL, value = "sample") %>%
  mutate(treatment = case_when(
    str_detect(sample, "CCl4") ~ "CCl4",
    str_detect(sample, "LPS") ~ "LPS",
    str_detect(sample, str_c(c("KD1", "KFHD"), collapse = "|")) ~ "control"
    )) %>%
  mutate(origin = case_when(str_detect(sample, "FHD") ~ "HC",
                            TRUE ~ "liver")) %>%
  mutate(time = case_when(str_detect(sample, "D1") ~ 1,
                          str_detect(sample, "D3") ~ 3)) %>%
  unite(group, origin, treatment, time, remove = F) %>%
  mutate(treatment = factor(treatment, levels = c("control","LPS", "CCl4")),
         origin = factor(origin, levels = c("liver", "HC")),
         time = ordered(time),
         group = as_factor(group)) 

# save meta data
saveRDS(meta, here(output_path, "meta_data.rds"))
```

## Exploratory analysis
### PCA of normalized data
PCA plot of normalized expression data contextualized based on the origin and treatment. Only the top 1000 most variable genes are used as features.
```{r pca-norm-data}
expr <- readRDS(here(output_path, "normalized_expression.rds"))
meta <- readRDS(here(output_path, "meta_data.rds"))

pca_result <- do_pca(expr, meta, top_n_var_genes = 1000)

saveRDS(pca_result, here(output_path, "pca_result.rds"))

plot_pca(pca_result, feature = "origin") +
  plot_pca(pca_result, feature = "treatment") &
  my_theme()
```

## Differential gene expression analysis
### Running limma
Differential gene expression analysis via limma with the aim to identify the effect of LPS intoxication.
```{r running-limma}
# load expression and meta data
expr <- readRDS(here(output_path, "normalized_expression.rds"))
meta <- readRDS(here(output_path, "meta_data.rds"))

stopifnot(colnames(expr) == meta$sample)

# build design matrix
design = model.matrix(~0+group, data=meta)
rownames(design) = meta$sample
colnames(design) = levels(meta$group)


# define contrasts
contrasts = makeContrasts(
  # LPS vs control in liver and hepatocytes
  inLiver_lps_vs_ctrl = liver_LPS_1 - liver_control_1,
  inHC_lps_vs_ctrl_day1 = HC_LPS_1 - HC_control_1,
  
  # CCl4 vs control in hepatocytes for day 1 and 3
  inHC_ccl_vs_ctrl_day1 = HC_CCl4_1 - HC_control_1,
  inHC_ccl_vs_ctrl_day3 = HC_CCl4_3 - HC_control_1,
  
  # LPS vs CCl4 in hepatocytes day 1 and 3
  inHC_ccl_vs_lps_day1 = HC_CCl4_1 - HC_LPS_1,
  inHC_ccl_vs_lps_day3 = HC_CCl4_3 - HC_LPS_1,
  
  # liver tissue vs hepatocytes
  liver_vs_hc_lps = liver_LPS_1 - HC_LPS_1,
  
  hc_vs_liver_ctrl = HC_control_1 - liver_control_1,
  levels = design
)

limma_result = run_limma(expr, design, contrasts) %>%
  assign_deg()

saveRDS(limma_result, here(output_path, "limma_result.rds"))
```

### Volcano plots
Volcano plots visualizing the effect of LPS on gene expression.
```{r volcano-plots}
df <- readRDS(here(output_path, "limma_result.rds"))

df %>%
  plot_volcano() + 
  my_theme(grid = "y", fsize = fz)
```
