---
title: "Chronic CCl4 (Fig. 1)"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r chunk-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  autodep = TRUE,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction
Here we generate publication ready plots of the analysis of the chronic CCl4 mouse model.

## Libraries and sources
These libraries and sources are used for this analysis.
```{r libs-and-src, message=FALSE, warning=FALSE, cache=FALSE}
library(tidyverse)
library(tidylog)
library(here)

library(AachenColorPalette)
library(cowplot)
library(lemon)
library(ggpubr)
library(VennDiagram)
library(grid)
library(gridExtra)
library(patchwork)

source(here("code/utils-plots.R"))
```

Definition of global variables that are used throughout this analysis.
```{r analysis-specific-params, cache=FALSE}
# i/o
data_path <- "data/mouse-chronic-ccl4"
output_path <- "output/mouse-chronic-ccl4"

# graphical parameters
# fontsize
fz <- 7

# keys to annotate contrasts
key_mm <- readRDS(here("data/meta-chronic-vs-acute/contrast_annotation.rds"))
```

## Design
```{r design}
design <- ggdraw() +
  draw_image(here(data_path, "exp-design.pdf"))

design
```

## Histology
```{r histology}
histology <- ggdraw() +
  draw_image(here(data_path, "histology.png"))

histology
```

## Liver enyzmes
```{r liver-enzymes}
df <- read_csv2(here(data_path, "liver_enzymes.csv")) %>%
  mutate(time = fct_inorder(time)) %>%
  pivot_longer(col = -c(time), names_to = "enzyme", values_to = "y") %>%
  mutate(enzyme = factor(str_to_upper(enzyme), levels = c("ALT", "AST", "ALP")))

df_summary <- df %>%
  group_by(time, enzyme) %>%
  summarise(mean_se(y)) %>%
  ungroup()

liver_enzymes <- df_summary %>%
  ggplot(aes(x = time, y = y)) +
  geom_errorbar(aes(ymin = ymin, ymax = ymax), width = 0.5) +
  geom_col() +
  facet_rep_wrap(~enzyme, scales = "free", ncol = 1) +
  labs(x = NULL, y = "U/L") +
  my_theme(grid = "y", fsize = fz) +
  stat_compare_means(
    data = df, label = "p.signif",
    ref.group = "Control", hide.ns = T
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

liver_enzymes
```

## PCA
```{r pca}
pca_result <- readRDS(here(output_path, "pca_result.rds"))

keys <- key_mm %>%
  filter(treatment == "CCl4" & class == "Chronic") %>%
  distinct(time = value, label = time_label2) %>%
  drop_na() %>%
  add_row(time = 0, label = "Control") %>%
  mutate(time = ordered(time))

pca_plot <- pca_result$coords %>%
  arrange(time) %>%
  inner_join(keys, by = "time") %>%
  mutate(label = fct_inorder(label)) %>%
  mutate(treatment = case_when(treatment == "oil" ~ "Oil",
                               treatment == "ccl4" ~ "CCl4 + Oil",
                               treatment == "ctrl" ~ "Control")) %>%
  ggplot(aes(x=PC1, y=PC2, color=label, shape = treatment, label = label)) +
  geom_point() +
  labs(x = paste0("PC1", " (", pca_result$var[1], "%)"),
       y = paste0("PC2", " (", pca_result$var[2], "%)"),
       color = "Time", shape = "Treatment") +
  my_theme(fsize  = fz) +
  theme(legend.position = "top",
        legend.box.margin = margin(10, 0, -20, 10)) +
  scale_color_manual(values = aachen_color(c(
    "violet", "bordeaux", "red", "orange"
  )))

pca_plot
```

## Volcano plot
```{r volcano-plot}
df <- readRDS(here(output_path, "limma_result.rds")) %>%
  filter(contrast_reference == "pure_ccl4") %>%
  inner_join(key_mm, by = "contrast") %>%
  select(-contrast) %>%
  rename(contrast = time_label2) %>%
  mutate(contrast = fct_drop(contrast)) %>%
  mutate(regulation = fct_recode(regulation,
    Up = "up", Down = "down",
    n.s. = "ns"
  ))

deg_count <- df %>%
  add_count(contrast, regulation) %>%
  filter(regulation != "n.s.") %>%
  mutate(regulation = fct_drop(regulation)) %>%
  mutate(
    logFC = case_when(
      regulation == "Up" ~ 0.75 * max(logFC),
      regulation == "Down" ~ 0.75 * min(logFC)
    ),
    pval = 0.4
  ) %>%
  distinct(n, contrast, logFC, pval, regulation, value) %>%
  complete(contrast, nesting(regulation, logFC, pval), fill = list(n = 0))

volcano <- df %>%
  plot_volcano(ncol = 1) +
  geom_text(
    data = filter(deg_count),
    aes(y = pval, label = n), size = fz / (14 / 5),
    hjust = "inward", vjust = "inward",
    show.legend = F
  ) +
  theme(
    legend.position = "top",
    legend.box.margin = margin(-10, 0, -10, 0)
  ) +
  labs(color = "Regulation") +
  my_theme(grid = "y", fsize = fz)

volcano
```

## Top DEGs
```{r top-degs}
df <- readRDS(here(output_path, "limma_result.rds")) %>%
  filter(contrast_reference == "pure_ccl4") %>%
  inner_join(key_mm) %>%
  select(-class) %>%
  rename(class = time_label2)

top_genes_df <- df %>% 
  group_by(class, sign(logFC)) %>%
  slice_max(order_by = abs(statistic), n = 10, with_ties = F) %>%
  ungroup() %>% 
  nest(data = -c(class, value))

plots <- top_genes_df %>%
  mutate(p = pmap(., .f = plot_top_genes, fontsize = fz))

# for main panel
top_genes <- plots %>%
  pull(p) %>%
  wrap_plots() +
  plot_layout(ncol = 1)

top_genes
```


## Gene overlap
```{r gene-overlap}
df <- readRDS(here(output_path, "limma_result.rds"))

tables = df %>%
  filter(contrast_reference == "pure_ccl4") %>%
  mutate(class = str_c("Month ", parse_number(as.character(contrast)))) %>%
  select(-contrast_reference, -contrast) %>%
  mutate(class = factor(class, levels = c("Month 2", "Month 6", "Month 12"))) %>%
  group_split(class)

# extract labellers
c1 = tables[[1]] %>% distinct(class) %>% pull() %>% as.character()
c2 = tables[[2]] %>% distinct(class) %>% pull() %>% as.character()
c3 = tables[[3]] %>% distinct(class) %>% pull() %>% as.character()

t1 = tables[[1]] %>% count(regulation)
t2 = tables[[2]] %>% count(regulation)
t3 = tables[[3]] %>% count(regulation)
    
plots = c("up", "down") %>%
  map(function(r) {
    # set sizes of regulated genes
    a1 = t1 %>% filter(regulation == r) %>% pull(n)
    a2 = t2 %>% filter(regulation == r) %>% pull(n)
    a3 = t3 %>% filter(regulation == r) %>% pull(n)
    
    a12 = purrr::reduce(
      list(tables[[1]] %>% filter(regulation == r) %>% pull(gene),
           tables[[2]] %>% filter(regulation == r) %>% pull(gene)),
      intersect) %>%
      length()
    
    a23 = purrr::reduce(
      list(tables[[2]] %>% filter(regulation == r) %>% pull(gene),
           tables[[3]] %>% filter(regulation == r) %>% pull(gene)),
      intersect) %>% 
      length()
    
    a13 = purrr::reduce(
      list(tables[[1]] %>% filter(regulation == r) %>% pull(gene),
           tables[[3]] %>% filter(regulation == r) %>% pull(gene)),
      intersect) %>% 
      length()
    
    a123 = purrr::reduce(
      list(tables[[1]] %>% filter(regulation == r) %>% pull(gene),
           tables[[2]] %>% filter(regulation == r) %>% pull(gene),
           tables[[3]] %>% filter(regulation == r) %>% pull(gene)),
      intersect) %>% 
      length()
    

    
    grid.newpage()
    p = draw.triple.venn(
      area1 = a1, area2 = a2, area3 = a3, 
      n12 = a12, n23 = a23, n13 = a13, 
      n123 = a123,
      category = c(c1, c2, c3),
      # lty = "blank",
      cex = 1/12*fz,
      fontfamily = rep("sans", 7),
      # fill = aachen_color(c("purple", "petrol", "red")),
      # cat.col = aachen_color(c("purple", "petrol", "red")),
      cat.cex = 1/12*(fz+1),
      cat.fontfamily = rep("sans", 3),
      cat.pos = c(0,0,180),
      cat.prompts = T
    ) %>%
      as_ggplot() %>%
      grid.arrange(top = textGrob(str_to_title(r), gp=gpar(fontsize=fz, 
                                             fontface = "bold")))
  })

gene_overlap = wrap_plots(plots)

```

## Top gene of overlap
```{r top-genes-overlap}
df = readRDS(here(output_path, "limma_result.rds")) %>%
  filter(contrast_reference == "pure_ccl4") %>%
  inner_join(key_mm) %>%
  select(-class) %>%
  rename(class = time_label2) %>% 
  filter(regulation != "ns") %>%
  mutate(regulation = fct_inorder(str_to_title(regulation)))
  
top_genes_ranked = df %>%
  # filter for genes that are deregulated at all time points
  group_by(gene, regulation) %>%
  filter(n() == 3) %>%
  summarise(mean_logfc = mean(logFC)) %>%
  group_by(regulation) %>%
  mutate(rank = row_number(-abs(mean_logfc))) %>%
  ungroup()

top_genes_of_overlap = df %>%
  inner_join(top_genes_ranked, by=c("gene", "regulation"))

top_overlap_genes = top_genes_of_overlap %>%
  filter(rank <= 5) %>%
  ggplot(aes(x=fct_reorder(gene, mean_logfc), y=logFC, group = class, 
             fill = class)) +
  geom_col(position = "dodge") +
  facet_rep_wrap(~regulation, ncol = 1, scales = "free") +
  my_theme(grid = "y", fsize = fz) +
  labs(x="Gene", y="logFC", fill = NULL) +
  scale_fill_manual(values = aachen_color(c("maygreen", "green", "turquoise")))

top_overlap_genes
```

## Time series cluster
```{r ts-cluster}
profile_label = tribble(
  ~profile, ~process,
  "STEM ID: 14", "Inflammation",
  "STEM ID: 6", "Proliferation",
  "STEM ID: 17", NA_character_,
  "STEM ID: 12", NA_character_,
  "STEM ID: 7", "Metabolism (1)",
  "STEM ID: 9", "Metabolism (2)",
  "STEM ID: 13", "ECM"
) %>%
  mutate(profile = fct_inorder(profile),
         process = coalesce(process, profile),
         process = fct_inorder(process))

stem_res <- readRDS(here(output_path, "stem_result.rds")) %>%
  filter(key == "pure_ccl4") %>%
  filter(p <= 0.05) %>%
  mutate(profile = fct_reorder(str_c(
    "STEM ID: ",
    as.character(profile)
  ), p)) %>%
  inner_join(profile_label, by = "profile") %>%
  select(-profile) %>%
  rename(profile = process)

# extract meta data of profiles
profile_anno <- stem_res %>%
  group_by(key, profile, p) %>%
  mutate(y = 1.2 * abs(max(value))) %>%
  ungroup() %>%
  mutate(max_time = max(time)) %>%
  distinct(key, profile, p, size, y, max_time) %>%
  mutate(label = str_c(size, " ", "genes"))

ts_cluster <- stem_res %>%
  plot_stem_profiles(model_profile = F, nrow = 2) +
  labs(x = "Time in Months") +
  geom_text(
    data = profile_anno, aes(x = 0, y = y, label = label),
    inherit.aes = F, size = fz / (14 / 5), hjust = "inward"
  ) +
  my_theme(grid = "no", fsize = fz) +
  scale_x_continuous(
    breaks = unique(stem_res$time),
    guide = guide_axis(n.dodge = 1)
  )

ts_cluster
```

## Collage
### Figure 1
```{r fig-1, fig.width=7.87, fig.height=11.69}
fig1 <- (design + histology) /
  ((liver_enzymes | gene_overlap | top_overlap_genes) + 
     plot_layout(widths = c(1,2,2))) /
  ts_cluster +
  plot_layout(height = c(1, 1, 1)) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "D", "", "E", "F"))) &
  theme(
    plot.tag = element_text(size = fz + 3, face = "bold"),
    legend.key.height = unit(11.5, "pt"),
    legend.key.width = unit(12.5, "pt")
  )

fig1

ggsave(here("figures/Figure 1.pdf"), fig1,
  width = 21, height = 29.7, units = c("cm")
)
ggsave(here("figures/Figure 1.png"), fig1,
  width = 21, height = 29.7, units = c("cm")
)
```

### Supplementary Figure 1.1
```{r s-fig-1-1, fig.width=7.87, fig.height=11.69}
sfig1_1 <- pca_plot / (volcano | top_genes) +
  plot_annotation(tag_levels = list(c("A", "B", "C"))) &
  theme(
    plot.tag = element_text(size = fz + 3, face = "bold"),
    legend.key.height = unit(11.5, "pt"),
    legend.key.width = unit(12.5, "pt")
  )

sfig1_1

ggsave(here("figures/Supplementary Figure 1.1.pdf"), sfig1_1,
  width = 21, height = 29.7, units = c("cm")
)
ggsave(here("figures/Supplementary Figure 1.1.png"), sfig1_1,
  width = 21, height = 29.7, units = c("cm")
)
```
