---
title: "Precision and recall of chronic mouse models (Fig. 5)"
output: 
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

```{r chunk-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  autodep = TRUE,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r wall-time-start, cache=FALSE, include=FALSE}
# Track time spent on performing this analysis
start_time <- Sys.time()
```

## Introduction
Here we generate publication-ready plots for the integration of further chronic mouse models.

## Libraries and sources
These libraries and sources are used for this analysis.
```{r libs-and-src, message=FALSE, warning=FALSE, cache=FALSE}
library(tidyverse)
library(tidylog)
library(here)

library(AachenColorPalette)
library(scales)
library(UpSetR)
library(grid)
library(lemon)
library(ggrepel)
library(magick)
library(patchwork)

library(gtools)

source(here("code/utils-plots.R"))
```

Definition of global variables and functions that are used throughout this analysis.
```{r analysis-specific-params, cache=FALSE}
# i/o
data_path <- "data/meta-mouse-vs-human"
output_path <- "output/meta-mouse-vs-human"

# graphical parameters
# fontsize
fz <- 7

# keys to annotate contrasts
key_mm <- readRDS(here("data/meta-chronic-vs-acute/contrast_annotation.rds"))
key_hs <- readRDS(here("data/meta-mouse-vs-human/contrast_annotation.rds"))
```

## Chronic mouse model genes in number
```{r vis-chronic-mm-gene-numbers}
keys <- key_mm %>%
  distinct(study = contrast, label2)

df <- readRDS(here(output_path, "chronic_mouse_deg_numbers.rds")) %>%
  left_join(keys, by = "study") %>%
  mutate(label2 = coalesce(label2, study)) %>%
  mutate(regulation = fct_rev(str_to_title(regulation)))

num_genes <- df %>%
  ggplot(aes(y = fct_reorder(label2, n, sum), x = n, fill = regulation)) +
  geom_col(position = "dodge") +
  labs(
    x = "Number of differentially expressed genes",
    y = "Chronic mouse model",
    fill = "Regulation"
  ) +
  scale_x_log10(
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", scales::math_format(10^.x))
  ) +
  my_theme(grid = "x", fsize = fz) +
  scale_fill_manual(values = aachen_color(c("red75", "blue75"))) +
  annotation_logticks(sides = "b")

num_genes
```

## Similarity of chronic mouse models and patient cohorts
```{r similarity-chronic-mm-and-hs}
contrast_keys_human <- key_hs %>%
  distinct(contrast, label, source, phenotype) %>%
  unite(contrast, source, phenotype, contrast, sep = "-")

contrast_keys_mouse <- key_mm %>%
  filter(str_detect(contrast, "pure")) %>%
  distinct(contrast, label)

keys <- bind_rows(contrast_keys_mouse, contrast_keys_human)


o <- readRDS(here(output_path, "cross_species_similarity.rds"))

oo <- o %>%
  rename(contrast = set1) %>%
  left_join(keys, by = "contrast") %>%
  mutate(label = coalesce(label, contrast)) %>%
  rename(set1 = label, contrast = set2, contrast_set1 = contrast) %>%
  left_join(keys, by = "contrast") %>%
  mutate(label = coalesce(label, contrast)) %>%
  rename(set2 = label, contrast_set2 = contrast) %>%
  mutate(
    contrast_set1 = factor(contrast_set1, levels = levels(o$set1)),
    contrast_set2 = factor(contrast_set2, levels = levels(o$set2)),
    set1 = fct_reorder(set1, as.numeric(contrast_set1)),
    set2 = fct_reorder(set2, as.numeric(contrast_set2))
  ) %>%
  select(set1, set2, similarity)

gs_sim <- oo %>%
  # filter(str_detect(set2, "stage|ctrl|mild|early")) %>%
  ggplot(aes(
    x = set1, y = set2, fill = similarity,
    label = round(similarity, 3)
  )) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = aachen_color("green")) +
  labs(x = NULL, y = NULL, fill = "Overlap\ncoefficient") +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.line = element_blank(),
    axis.ticks = element_blank()
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  my_theme(fsize = fz, grid = "no")

gs_sim
```

## Enrichment of chronic mouse genes in patient cohorts
```{r enrichment-chronic-mm-in-hs}
contrast_keys_human <- key_hs %>%
  distinct(contrast, label, source) %>%
  unite(contrast, source, contrast)

contrast_keys_mouse <- key_mm %>%
  filter(str_detect(contrast, "pure")) %>%
  distinct(contrast, label)

keys <- bind_rows(contrast_keys_mouse, contrast_keys_human)

gsea_res <- readRDS(here(output_path, "cross_species_enrichment.rds")) %>%
  rename(contrast = signature) %>%
  inner_join(keys, by = "contrast") %>%
  select(-contrast) %>%
  rename(signature = label, contrast = geneset) %>%
  left_join(keys, by = "contrast") %>%
  mutate(label = coalesce(label, contrast)) %>%
  select(-contrast) %>%
  rename(geneset = label) %>%
  mutate(direction = fct_rev(str_to_title(direction)))

tile_up <- gsea_res %>%
  filter(direction == "Up") %>%
  mutate(
    label = stars.pval(padj),
    direction = fct_rev(direction)
  ) %>%
  ggplot(aes(x = signature, y = geneset, fill = ES)) +
  geom_tile() +
  geom_text(aes(label = label)) +
  facet_rep_wrap(~direction, scales = "free", ncol = 1) +
  my_theme(grid = "no", fsize = fz) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.line = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient2(
    low = aachen_color("blue"), mid = "white",
    high = aachen_color("red"), limits = c(-1, 1)
  ) +
  my_theme(grid = "no", fsize = fz) +
  labs(x = "Signature", y = "Gene Set", fill = "ES") +
  guides(fill = guide_colorbar(title = "ES"))

box_up <- gsea_res %>%
  filter(direction == "Up") %>%
  ggplot(aes(y = geneset, x = ES)) +
  geom_boxplot() +
  facet_rep_wrap(~direction, ncol = 1) +
  geom_vline(xintercept = 0) +
  my_theme(fsize = fz, grid = "x") +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank()
  )

tile_down <- gsea_res %>%
  filter(direction == "Down") %>%
  mutate(
    label = stars.pval(padj),
    direction = fct_rev(direction)
  ) %>%
  ggplot(aes(x = signature, y = geneset, fill = ES)) +
  geom_tile() +
  geom_text(aes(label = label)) +
  facet_rep_wrap(~direction, scales = "free", ncol = 1) +
  my_theme(grid = "no", fsize = fz) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.line = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient2(
    low = aachen_color("blue"), mid = "white",
    high = aachen_color("red"), limits = c(-1, 1)
  ) +
  my_theme(grid = "no", fsize = fz) +
  labs(x = "Signature", y = "Gene Set", fill = "ES") +
  guides(fill = guide_colorbar(title = "ES"))

box_down <- gsea_res %>%
  filter(direction == "Down") %>%
  ggplot(aes(y = geneset, x = ES)) +
  geom_boxplot() +
  facet_rep_wrap(~direction, ncol = 1) +
  geom_vline(xintercept = 0) +
  my_theme(fsize = fz, grid = "x") +
  theme(
    axis.text.y = element_blank(),
    axis.title.y = element_blank(),
    axis.line.y = element_blank(),
    axis.ticks.y = element_blank()
  )

tile_up + box_up
tile_down + box_down
```

## Etiology genes in numbers
```{r vis-etiology-gene-numbers}
df <- readRDS(here(output_path, "etiology_gene_sets.rds"))

mat_up <- df %>%
  filter(regulation == "up") %>%
  select(-regulation) %>%
  mutate(val = 1) %>%
  spread(etiology, val, fill = 0) %>%
  data.frame(row.names = 1)

pdf(file = here("figures/tmp/Fig5A1.pdf"), width = 15, height = 10, onefile = F)
upset(mat_up,
  nintersects = NA, mainbar.y.label = "Common genes",
  sets.x.label = "Total number of genes",
  text.scale = 3, point.size = 5
)
grid.text("Up", x = 0.65, y = 0.95, gp = gpar(fontsize = fz * 3))
dev.off()

mat_down <- df %>%
  filter(regulation == "down") %>%
  select(-regulation) %>%
  mutate(val = 1) %>%
  spread(etiology, val, fill = 0) %>%
  data.frame(row.names = 1)

pdf(file = here("figures/tmp/Fig5A2.pdf"), width = 15, height = 10, onefile = F)
upset(mat_down,
  nintersects = NA, mainbar.y.label = "Common genes",
  sets.x.label = "Total number of genes",
  text.scale = 3, point.size = 5
)
grid.text("Down", x = 0.65, y = 0.95, gp = gpar(fontsize = fz * 3))
dev.off()
```

## PR scatter plot
```{r pr-scatter-plot}
keys <- key_mm %>%
  distinct(study = contrast, label2)

pr <- readRDS(here(output_path, "precision_recall.rds")) %>%
  left_join(keys, by = "study") %>%
  mutate(label2 = coalesce(label2, study)) %>%
  mutate(regulation = fct_rev(str_to_title(regulation))) %>%
  mutate(x = case_when(
    str_detect(label2, "12 Month") ~ as.character(label2),
    TRUE ~ NA_character_
  )) %>%
  mutate(etiology = factor(etiology, levels = c(
    "NAFLD", "NASH", "HCV", "PSC",
    "PBC"
  )))

pr_plot <- pr %>%
  ggplot(aes(x = recall, y = precision, label = x, color = label2)) +
  geom_point() +
  facet_rep_grid(etiology ~ regulation) +
  geom_text_repel(size = fz / (14 / 5), show.legend = FALSE, na.rm = TRUE) +
  geom_abline(lty = "dashed") +
  expand_limits(x = 0, y = 0) +
  labs(x = "Recall", y = "Precision", color = "Mouse model") +
  my_theme(fsize = fz) +
  scale_color_viridis_d() +
  coord_fixed()

pr_plot
```

## Collage
### Figure 5
Main Figure.
```{r fig-5, fig.width=8.27, fig.height=11.69}
upset_up <- image_ggplot(image_read_pdf(here("figures/tmp/Fig5A1.pdf")),
                         interpolate = TRUE)

# file.remove(here("figures/tmp/Fig5A1.pdf"))
upset_down <- image_ggplot(image_read_pdf(here("figures/tmp/Fig5A2.pdf")),
  interpolate = TRUE
)
# file.remove(here("figures/tmp/Fig5A2.pdf"))

fig5 <- (upset_up | upset_down) /
  ((pr_plot | plot_spacer()) + plot_layout(widths = c(2, 1))) +
  plot_layout(height = c(1, 3)) +
  plot_annotation(tag_levels = list(c("A", "", "B"))) &
  theme(
    plot.tag = element_text(size = fz + 3, face = "bold"),
    legend.key.height = unit(11.5, "pt"),
    legend.key.width = unit(12.5, "pt")
  )

fig5

ggsave(here("figures/Figure 5.pdf"), fig5,
  width = 21, height = 29.7, units = c("cm")
)
ggsave(here("figures/Figure 5.png"), fig5,
  width = 21, height = 29.7, units = c("cm")
)
```

### Supplementary Figure 5.1
```{r fig-5-1, fig.width=8.27, fig.height=11.69}
sfig5_1 <- ((num_genes | gs_sim) + plot_layout(width = c(1, 3))) /
  ((tile_up + box_up) + plot_layout(width = c(4, 1), guides = "collect")) /
  ((tile_down + box_down) + plot_layout(width = c(4, 1), guides = "collect")) +
  plot_annotation(tag_levels = list(c("A", "B", "C", "", "D"))) &
  theme(
    plot.tag = element_text(size = fz + 3, face = "bold"),
    legend.key.height = unit(11.5, "pt"),
    legend.key.width = unit(12.5, "pt")
  )

sfig5_1

ggsave(here("figures/Supplementary Figure 5.1.pdf"), sfig5_1,
  width = 21, height = 29.7, units = c("cm")
)
ggsave(here("figures/Supplementary Figure 5.1.png"), sfig5_1,
  width = 21, height = 29.7, units = c("cm")
)
```

```{r wall-time-end, cache=FALSE, include=FALSE}
duration <- abs(as.numeric(difftime(Sys.time(), start_time, units = "secs")))
t <- print(sprintf("%02d:%02d", duration %% 3600 %/% 60, duration %% 60 %/% 1))
```
Time spend to execute this analysis: `r t` minutes.
